---
title:  |
  **Group 2**  
  **P-MHD: Multivariate Methods**  
subtitle: |
  **Aaryan Kaushik (2159244), Deo Byabazaire (2159254), Edmond Sacla Aide (2159278),  Muhammad Bergas Nur  Fayyad (2159123), Daniel R. Yildirim (2056569), Wouter Smeets (1849125)**
date: "May 23, 2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
fontsize: 14pt
mainfont: Arial
header-includes: \usepackage{amsmath}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,	
                      warning = FALSE,
                      eval = TRUE,
                      tidy = FALSE)
library(magrittr)
library(dplyr)
library(ggpubr)
library(tidyverse)
library(GrassmannOptim)
library(Matrix)
library(gridExtra)
library(ggplot2)
library(maps)
library(ggrepel)
library(ldr)
library(bookdown)
```

# 1.Introduction

Precipitation and temperature are among the most important
elements to describe the climate in a certain area. This study aims to find
similarities and dissimilarities among Canadian
cities regarding the precipitation regime.

## 1.1 Data

The precipitation is the main data available in the Canadian weather
dataset. It contains the average daily rainfall (mm/day) for the 365
days in the year and for 35 Canadian cities. 
In addition, this dataset includes other information such as the regions,
provinces and coordinates. We have added a new column indicating the season for later use.

```{r ,echo=FALSE, include=FALSE}
load("CanadianWeather.rda")
da<-CanadianWeather[[1]]
da<-da[,,"Precipitation.mm"]
PD<-data.frame(da) # precipitation data
PD <- cbind(date = 1:365, PD) # date column
rownames(PD) <- 1:nrow(PD)
colnames(PD[,-c(1,37)])
```

```{r ,echo=FALSE, include=FALSE}
PDplot<-PD

PDplot$season <- ifelse(PD$date >= 1 & PD$date <= 59, 'winter',         
                  ifelse(PD$date >=60 & PD$date <=151, 'spring',
                         ifelse(PD$date >=152 & PD$date <=243, 'summer', 
                                ifelse(PD$date >=244 & PD$date <=334, 'autumn',
                                       ifelse(PD$date >=335 & PD$date <=365, 'winter', "NA")))))

a<-ggplot(aes(x=date, y=Victoria, color=season), data = PDplot) +
  geom_point() +
  labs(title="Victoria",
        x ="days", y = "precipitation (mm/day)")
  
b<-ggplot(aes(x=date, y=Toronto, color=season), data = PDplot) +
  geom_point() +
  labs(title="Toronto",
        x ="days", y = "precipitation (mm/day)")

grid.arrange(a, b, ncol=1)
```

## 1.2 Research question

The main objective of this project is to sort out the variability between
Canadian cities in terms of the average daily rainfall. It addresses
the questions of which cities have similar precipitation patterns, 
and those with dissimilar patterns? Also, how can the differences between
the cities be described (i.e., in what sense do cities differ)?

# 2.Functional Data Analysis (FDA)

## 2.1 Introduction

The Multi-Dimensional Scaling (MDS) is one of the multivariate methods
that aims to find a low-dimensional representation, say $k$ dimensional space, of $n$ data points such that the distances between the $n$ points in the $k$-dimensional space are a good approximation of a given squared
distance matrix, say $D_x$. Hence, the research question leads us to
consider it an appropriate method to address the objective of this
project.

## 2.2 Transformation to functions

The precipitation dataset used in this study consists of $n=35$ rows (cities) and
$p=365$ columns (days). This is high dimensional data (as $p>>n$).
This structure makes it easier to look at the raw data table. A
functional data analysis (FDA), which assumes that functions are
considered as observations, has been used to analyze this data. For each
city, the observations of the precipitation are in the function of
days.

To take this approach,
it is necessary to first convert the data entries for each cities to a single function. As the aim of Singular Value Decomposition (SVD) is to transform the dataset to a lower-dimensional dataset called  the parameter space, it therefore gives a matrix where $q<p$
is less than the original data. Consequently,
each city will have its set of $q$ parameter estimates, and so an
$(n\times q)$ data matrix can be constructed. To give a meaningful
interpretation to the results at the end of our analysis, the solution needed to be back-transformed from the parameter space to the functional space.

## 2.3 Multidimensional Scaling of Functions

Denote $Y_i(t)$ the outcome of observation $i=1,\ldots, n$ as the average daily
precipitation for cities $i$ at time $t$ between January 1 and December
31. For observation $i$ there are data on times $t_{ij}$,
$j=1,\ldots,p_i$.

Consider the non-linear model $$
  Y_i(t_{ij}) = f_i(t_{ij}) + \varepsilon_{ij}=\sum_{k=0}^m \theta_{ik} \phi_k(t_{ij}) + \varepsilon_{ij}\approx \sum_{k=0}^m \theta_{ik} x_{ijk} + \varepsilon_{ij};
  \;\; i=1,\ldots, n; j=1,\ldots, p_i; k=0,\ldots, m
  $$ where $f_i(\cdot)$ is a smooth function and $\varepsilon_{ij}$
i.i.d. with mean 0 and constant variance $\sigma^2$. The $\phi_k(\cdot)$
form a set of orthonormal basis functions.

The days were re-scaled to [0,1] interval to avoid numerical problems.

```{r ,echo=FALSE, include=FALSE}
days<-1:365
days<-(days-min(days))/(diff(range(days))) # re-scaling to [0,1]
```

Next basis functions, with either a polynomial basis or Fourier basis were compared such that the one that gave a better fit to the dataset would be chosen. Thus, a small simulation was done to compare the goodness of fit of the linear regression using the polynomial basis with the one using Fourier basis.
The comparison was based on the adjusted R-squared (adj-$R^2$) and the
mean square error (MSE).

```{r ,echo=FALSE, include=FALSE}
########### Which basis function will be used
phi1<-poly(days,degree=17)# Polynomial
phi2<-bf(days,case="fourier",degree=17) # fourrier basis
n<-length(colnames(PD))
Model_eval<-matrix(nrow = n, ncol = 4)
for(i in 1:n){
  model1<-lm(PD[,i]~phi1)
  model2<-lm(PD[,i]~phi2)
  Model_eval[i,1]<-summary(model1)$adj.r.squared
  Model_eval[i,2]<-summary(model2)$adj.r.squared
  
  Model_eval[i,3]<-mean(model1$residuals^2)
  Model_eval[i,4]<-mean(model2$residuals^2)
  Model_eval<-`rownames<-`(Model_eval,colnames(PD))
  Model_eval<-`colnames<-`(Model_eval,c("adj-R^2 Polynomial", " adj-R^2 Fourier", "MSE Polynomial", "MSE Fourier"))
}
head(Model_eval, 10)
```

The Fourier basis seemed to have a better prediction power compared to
the polynomial basis. Also from some previous studies by Tsai et al. (2016) & Adams et al. (2018) it was reported that the Fourier basis is often a better choice than the polynomial basis. 

Next the degree $m$ of the Fourier basis has been selected. For a given
city, different values of $m$ (1, 5, 10, 15,
16, 17, 18, 19, 20) were applied. It was observed that from $m=17$
and above, there was no additional variation in the
precipitation trend (see code). Hence, for the construction of the theta
matrix, Fourier basis with degree 17 was used.

```{r ,echo=FALSE, include=FALSE}
################# Which degree will be use
par(mfrow=c(3,3))
days<-1:365
days<-(days-min(days))/(diff(range(days))) # rescaling to [0,1]

# m=1
phi<-bf(days,case="fourier",degree=1)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=1)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)

# m=5
phi<-bf(days,case="fourier",degree=5)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=5)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)

# m=10
phi<-bf(days,case="fourier",degree=10)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=10)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)

# m=15
phi<-bf(days,case="fourier",degree=15)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=15)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)

# m=16
phi<-bf(days,case="fourier",degree=16)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=15)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)

# m=17
phi<-bf(days,case="fourier",degree=17)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=17)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)

# m=18
phi<-bf(days,case="fourier",degree=18)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=18)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)

# m=19
phi<-bf(days,case="fourier",degree=19)
dim(phi)
# estimation of the theta parameters for Vancouver
m.Vancouver<-lm(PD[,"Vancouver"]~phi)
summary(m.Vancouver)
# plot of fitted function
plot(1:365,PD[,"Vancouver"],main="Vancouver (m=18)", xlab="day", ylab="precipitation (mm/day)")
lines(1:365,m.Vancouver$fitted.values,type="l", col=2)
```

The estimation of the theta parameters for the chosen 35
cities was made.

```{r ,echo=FALSE, include=FALSE}
# now on log-transformed data
#PD[PD == 0] <- 0.05 # replacing zero values with 0.05
#PDlog[,-1]<-log(PDlog[,-1]+1)

# we consider 20 basis functions
days<-PD$date
day.first<-min(days)
day.last<-max(days)
days.range<-diff(range(days))

days.std<-(days-day.first)/(days.range) # rescaling to [0,1]
```

```{r ,echo=FALSE, include=FALSE}
# Create theta matrix using the polynomial basis functions up to degree 20
PD2<-PD[,-1]
n.cities<-35
phi<-bf(days.std,case="fourier",degree=17)

Theta<-matrix(nrow=n.cities,ncol=ncol(phi)+1)
for(i in 1:n.cities) {
  m<-lm(PD2[,i]~phi)
  Theta[i,]<-coef(m)
}
dim(Theta)
```

Finally, the statistical model for country $i$ in matrix
notation is as follows:. 
  $\boldsymbol{Y}_i = \boldsymbol{\theta}_i^t\boldsymbol{X}_i+\boldsymbol{\varepsilon}_i$ for $i = 1,...,35$

where $\boldsymbol{Y}_i$ is the vector with the outcomes of
observation $i$ (one for each day $t{ij}$), $\boldsymbol{\theta}_i$ the
vector with the $\theta_{ik}$ (one for each basis function $k$),
$\boldsymbol{X}_i$ the matrix with the $x{ijk}$ (days $j$ in the rows,
basis function index $k$ in columns), and $\boldsymbol{\varepsilon}_i$
the vector with the i.i.d. error terms. 

The parameters $\theta_{ik}$ can be estimated by means of least squares.

## 2.3 Multidimensional Scaling of Functions


Define $\hat{\boldsymbol{\theta}}_i$ as the vector with the
parameter estimates.
Then, the estimates for all cities can be collected into a single new
$n \times (m+1)$ data matrix $\boldsymbol{\Theta}$; where the $i^{th}$
row of $\boldsymbol{\Theta}$ is $\hat{\boldsymbol{\theta}}_i^t$. Since
$\boldsymbol{\Theta}$ has the structure of an ordinal data matrix, the
MDS was applied to $\boldsymbol{\Theta}$, so that a 
2-dimensional plot can be constructed with each point representing a Canadian's city. The
distances between the points in the 2-dimensional MDS space are
approximations of the distances between the rows of
$\boldsymbol{\Theta}$. Hence, it can be interpreted as distances between
the precipitation functions.

The MDS starts from the truncated SVD of $\boldsymbol{\Theta}$ (or
after column-centering), $$
   \boldsymbol{\Theta}_k = \boldsymbol{U}_k\boldsymbol{D}_k\boldsymbol{V}_k^t
$$ (note that the index $k$ now refers to the number of components in
the truncated SVD and not to the index of the basis functions).

Then the column centering and SVD were applied on the theta matrix.

```{r echo=TRUE, fig.show='hide', message=FALSE, include=FALSE}
Theta.mean<-colMeans(Theta)
Theta<-scale(Theta,center = TRUE, scale = FALSE)
```

```{r,fig.cap = "Squared Singular Value vs Index", fig.height = 4, fig.width = 6, fig.align = "center",echo=FALSE, warning=FALSE}
Theta.svd<-svd(Theta)
plot(Theta.svd$d^2/(sum(Theta.svd$d^2)),type="l",
     ylab="squared singular value")
```
Figure 1 indicates that the first few dimensions capture most of the information in the $\Theta$ matrix.
We can see a rapid decrease of the singular values from the larger values. This implies the information in terms of variability contained in the squared singular values decreases rapidly and eventually tends to zero.

# 3.Functional biplot

The interpretation of the functional biplot is easier after transforming the SVD to
the original function space.

The fitted model for all $n$ cities may then be simultaneously written
as $$
  \hat{\boldsymbol{Y}} = \boldsymbol{\Theta} \boldsymbol{X}^t
$$ with $\hat{\boldsymbol{Y}}$ the $n \times p$ matrix with $i$th row
$\boldsymbol{Y}_i^t$, and $\boldsymbol{X}$ the $p \times (m+1)$ matrix
$\boldsymbol{X}_i$ as defined before (note that all $\boldsymbol{X}_i$
are equal because for all cities $i$ the measurements were obtained at
the same $p$ time points (years)).

After substituting $\boldsymbol{\Theta}$ with its truncated SVD (after
$k$ terms), the simplified model function can have the form: $$
  \hat{Y}_{ki}(t)=  \sum_{j=1}^k \sum_{r=0}^m z_{kij} v_{rj} \phi_r(t)
$$ where $z_{kij}$ is the $(i,j)$th element of $\boldsymbol{Z}_k$ and
$v{rj}$ is the $(r,j)$th element of $\boldsymbol{V}_k$

The following figure presents the functional-plot. The origin of the
graph for both dimensions corresponding to the average precipitation
function starts at (0,0). This is because $\boldsymbol{\Theta}$ is
column-centered. 
Figure 2 suggests that both in the first and second dimension, some cities have negative score and some have positive score.

In the first dimension, Pr Rupert, St
Johns, Sydney, Halifax, Yarmouth and Varcouver have higher negative
score. These cities are opposed to Resolute, Inuvik, Whitehorse, Regina,
Pr Albert and Calgary that present positive score.

In the second dimension, Victoria and Vancouver have negative score
whereas Quebec, Sherbrook, Thundi and Winnipeg seem to have significant
positive score.

```{r,fig.height = 4, fig.width = 7, fig.align = "center",echo=FALSE, include=FALSE}
## Fonctional plot
Z<-Theta.svd$u[,1:2]%*%diag(Theta.svd$d[1:2])
plot(Z[,1],Z[,2])
rownames(Z)<-rownames(Theta[,-1])
text(Z[,1],Z[,2],rownames(Z))
abline(v=0,lty=2,col=4)
abline(h=0,lty=2,col=4)
```

```{r,fig.cap = "Functional biplot of Precipitation of Canadian Cities",fig.height = 4, fig.width = 8, fig.align = "center",echo=FALSE, warning=FALSE}
## Functional biplot with province info
Z<-Theta.svd$u[,1:2]%*%diag(Theta.svd$d[1:2])
Z<-as.data.frame(Z)
Z$province<-CanadianWeather[["province"]]
Z$city<-colnames(PD[,-c(1,37)])
rownames(Z)<-1:nrow(Z)


ggplot(aes(x=V1, y=V2, color=province), data = Z) + geom_vline(xintercept = 0) + geom_hline(yintercept = 0) +
  geom_point(size = 0.5) + geom_label_repel(aes(label = city),
                  box.padding   = 0.15, 
                  point.padding = 0.2,
                  segment.color = 'grey50') +
  theme_classic() + 
  labs(title="Functional Biplot",
        x ="first dimension", y = "second dimension")
```
From the functional biplot (Figure 2), although there were no other important geographic and topological features of cities that influence precipitation included, such as latitude, proximity to large bodies of water, and location relative to mountain ranges, we see it can be seen that most cities in the same provinces have a similar rainfall regime.

For a further interpretation of the functional biplot, the SVD was transformed back to the original function space.

```{r,fig.height = 4, fig.width = 6,echo=FALSE, include=FALSE}
days.range<-365-1

# right signular vectors (loadings)
V<-Theta.svd$v[,1:2]

# a vector with many points in the [0,1] interval.
# this will be used for plotting (at each point I
# "predict" a precipitation function value)
days.std.plot<-seq(0,1,length.out = 365)

# evaluate the polynomial basis functions at all these points
phi.plot<-bf(days.std.plot,case="fourier",degree=17)

# construct the design matrix for the linear model
# (note that I need to add a column with ones,
# corresponding to the degree 0 polynomial)
X<-cbind(1,phi.plot)

# product of this X matrix with the V matrix
XV<-X%*%V

# X times the vector with the column means of the
# original Theta matrix gives an estimate of the
# average precipitation function (evaluated
# in the 365 points in the [0,1] interval)
# Note that this is a mean on the log scale!
mean.curve<-X%*%Theta.mean

# we plot the mean.curve on the original scale. However,
# this is not an unbiased estimator of the average
# precipitation function on its original scale. For our
# purposes, this does not matter. It serves as a kind of
# reference function.
plot(days.std.plot,mean.curve, type="l")
```

Based on Figure 3, it was possible to conclude that cities with a
large negative score (red line) have high and decreasing precipitation average from late winter to the middle of summer and increasing precipitation
average from mid summer until the end of autumn. Then cities with
negative scores have high precipitation compared to the average. The
average precipitation for cities with large positive scores (blue line) increases
from late winter until mid summer and decreases until early
winter. These cities with positive scores have low precipitation
compared to the average. Whereas, the black line represents average rainfall for all cities and shows similar rainfall patterns for all the seasons and there are no clear visible peaks.

```{r, fig.cap = "Back-transformed to the original function space in the First Dimension",fig.height = 4, fig.width = 6,echo=FALSE}
###  The scores in the 1st axis vary between -6 and 1.5
rng<-range(mean.curve-XV[,1]*6,mean.curve+XV[,1]*1.5)
plot(days.std.plot*days.range,mean.curve,type="l",ylim=rng,
     xlab="Days after 1st January", ylab="Precipitation (mm)")
abline(v=c(59,150,243,335), col=c("gray"), lty=c(2), lwd=c(1))
lines(days.std.plot*days.range,mean.curve+XV[,1]*1.5,col=4)
lines(days.std.plot*days.range,mean.curve-XV[,1]*6,col=2)
```

Figure 4 shows that cities with negative scores (red line) in the second dimension have a low precipitation average compared to the average from spring to mid autumn. In contrast, cities with a positive score in that dimension have higher precipitation than the average. Moreover, we observe an opposite trend for the rest of the year (mid autumn to winter).

```{r, fig.cap = "Back-transformed to the original function space in the Second dimension", fig.height = 4, fig.width = 6,echo=FALSE}
### The scores in the second axis vary between -1.5 and 1
rng<-range(mean.curve-XV[,2]*1.5,mean.curve+XV[,2]*1)
plot(days.std.plot*days.range,mean.curve,type="l",ylim=rng,
     xlab="Days after 1st January", ylab="Precipitation (mm)")
abline(v=c(59,150,243,335), col=c("gray"), lty=c(2), lwd=c(1))  #for seasons
lines(days.std.plot*days.range,mean.curve+XV[,2]*1,col=4)
lines(days.std.plot*days.range,mean.curve-XV[,2]*1.5,col=2)
```

```{r}

```
\pagebreak
# Conclusion

Considering the information from the score plot and  the trend plots i.e first dimension plot that includes Pr Rupert, St Johns, Sydney, Halifax, Yarmouth and Varcouver have a
daily precipitation higher than the average during the whole year. But
this precipitation decreases from late winter to the middle mid
summer and increases the rest of the year. In Resolute, Inuvik, Whitehorse, Regina,
Pr Albert, and Calgary, the daily precipitation is lower than the
average. It increases from late winter to mid summer and decreases
the rest of the year. In Victoria, the average precipitation is lower
than the average from spring to mid autumn but higher than the
average for the rest of the year. Furthermore, in Quebec, Sherbrook, Thundi
and Winnipeg, the daily precipitation is higher than the average from
spring to middle mid autumn.

# References

Tsai, Cho-Liang, Wei Tong Chen, and Chin-Shiang Chang. "Polynomial-Fourier series model for analyzing and predicting electricity consumption in buildings." Energy and Buildings 127 (2016): 301-312.

2022. Cs.Princeton.Edu. \url{https://www.cs.princeton.edu/courses/archive/fall18/cos324/files/basis-functions.pdf}.
